A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-04-29 submission/Highlights.rmd
add missing references

# Highlights
  - Standardized framework for risk-based assessment of treatment effect
    heterogeneity
  - Always run the framework after population-level effect estimation problems
  - Run federated analyses across different databases mapped to OMOP common data
    model

>>>>

# Highlights
  - Standardized framework for risk-based assessment of treatment effect
    heterogeneity
  - Always run the framework after population-level effect estimation problems
  - Enables federated analyses across databases mapped to OMOP common data model

### 2021-04-29 submission/manuscript.rmd
add missing references

email: a.rekkas@erasmusmc.nl
\onehalfspacing

@@

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observation Health Data Sciences and Informatics (OHDSI)
initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
methods library with a framework for risk-based assessment of treatment effect
heterogeneity. The proposed framework consists of five steps: 1) definition of
the problem, i.e. the population, the treatment, the comparator and the
outcome(s) of interest; 2) identification of relevant databases; 3) development
of a prediction model for the outcome(s) of interest; 4) estimation of
propensity scores within strata of predicted risk and estimation of relative and
absolute treatment effect within strata of predicted risk; 5) evaluation and
presentation of results. We demonstrate our framework by evaluating
heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. he proposed framework may be useful for the evaluation of heterogeneity
of treatment effect on observational data that are mapped to the OMOP Common
Data Model. The proof of concept study demonstrates its feasibility in large
observational data. Further insights may arise by application to safety and
effectiveness questions across the global data network.

@@

Interest in understanding how a treatment’s effect varies across patients—a
concept described as heterogeneity of treatment effects (HTE)—has been growing.
This concept is central to the agenda for both personalized (or precision)
medicine and comparative effectiveness research. More formally, HTE has been
defined as non-random variability in the direction or magnitude of a treatment
effect, in which the effect is measured using clinical outcomes
[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
effects are assessed one at a time in patient subgroups defined from single
covariates, an approach that suffers from low power and multiplicity issues
[@Varadhan2013]. However, even with well-established constant relative
effects, treatment benefit (or harm) may vary substantially on the absolute
scale.

@@

While these approaches have generally been recommended for application to
clinical trials, observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, and include heterogeneous populations. However, unlike trials,
treatment effects are subject to confounding and the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
@vanKlaveren2019]. To do this, we develop the patient-level prediction model in
the propensity score-matched subset of the population (1:1), where treatment
assignment is well-balanced. The propensity scores are based on LASSO logistic
regression for modeling the association between treatment assignment and all
available demographics, drug exposures, diagnoses, measurements and medical
procedures. Finally, we need to define the time horizon for which we aim to make
predictions and we need to select the machine-learning algorithm we want to use
to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

@@

The number of patients and person years by treatment arm along with the number
of outcomes should be reported. A performance overview of the derived prediction
models, including discrimination and calibration both in the propensity score
matched subset, the entire population and separately for treated and comparator
patients should also be reported. This is rather relevant as the performance of
the prediction models is directly related to our ability to single out patient
subgroups where treatment may be highly beneficial or unsafe. Kent et al
[@Kent2016] demonstrated that the event rate and the discriminative ability of
the prediction model can predict very well the distribution of predicted risk.
The lower the event rate and the higher the c-statistic (given good calibration)
result in high risk heterogeneity, thus making estimated average treatment
effects uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

In every risk stratum propensity score distributions by treatment group and
covariate balance plots should be made available along with event rates, hazard
ratios and absolute risk differences. Finally, the results should be generated
with a standardized format to allow for combination of results from multiple
data sources. The standardized result format also enables the development of
scalable demonstration tools such as R-shiny applications, interactive tables
and figures.

@@

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a rough guide for
clinical interpretation, it is not appropriate to guide clinical practice, where
decisions need to be made at the individual level [@Kent2019]. Presentation of
treatment effects as a continuous function of risk would be more helpful, but is
methodologically challenging. Future research is necessary for the development
of methods for continuous risk-based assessment of HTE.

Externally derived prediction models are preferred for analyzing treatment
effect heterogeneity [@Kent2010]. Such external models should be well
transportable. In the absence of such prediction models, simulations of RCTs
have shown that internally derived models can be used to provide unbiased
estimates of treatment effect across the spectrum of baseline risk
[@Burke2014]. However, in observational databases treatment arms may
significantly differ in sample size. Because the prediction model will possibly
better fit to the larger treatment arm, this may introduce spurious
treatment-covariate interactions in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.

>>>>

Email: a.rekkas@erasmusmc.nl
\onehalfspacing

@@

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics (OHDSI)
initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
methods library with a framework for risk-based assessment of treatment effect
heterogeneity. The proposed framework consists of five steps: 1) definition of
the problem, i.e. the population, the treatment, the comparator and the
outcome(s) of interest; 2) identification of relevant databases; 3) development
of a prediction model for the outcome(s) of interest; 4) estimation of
propensity scores within strata of predicted risk and estimation of relative and
absolute treatment effect within strata of predicted risk; 5) evaluation and
presentation of results. We demonstrate our framework by evaluating
heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of heterogeneity
of treatment effect on observational data that are mapped to the OMOP Common
Data Model. The proof of concept study demonstrates its feasibility in large
observational data. Further insights may arise by application to safety and
effectiveness questions across the global data network.

@@

Understanding how a treatment’s effect varies across patients—a concept
described as heterogeneity of treatment effects (HTE)—has been central to the
agenda for both personalized (or precision) medicine and comparative
effectiveness research. More formally, HTE has been defined as the non-random
variability in the direction or magnitude of a treatment effect, in which the
effect is measured using clinical outcomes [@KRAVITZ2004]. Usually, analyses
focus on the relative scale, where treatment effects are assessed one at a time
in patient subgroups defined from single covariates, an approach that suffers
from low power and multiplicity issues [@Yusuf1991; @FernandezyGarcia2010].
However, even with well-established constant relative effects, treatment benefit
(or harm) may vary substantially on the absolute scale.

@@

While these approaches have generally been recommended for application to
clinical trials, observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, and include heterogeneous populations. However, unlike trials,
treatment effects are subject to confounding while the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arms in order to avoid inducing spurious interactions
[@Burke2014; @vanKlaveren2019]. To do this, we develop the patient-level
prediction model in the propensity score-matched subset of the population (1:1),
where treatment assignment is well-balanced. The propensity scores are based on
LASSO logistic regression for modeling the association between treatment
assignment and all available demographics, drug exposures, diagnoses,
measurements and medical procedures. Finally, we need to define the time horizon
within which we aim to make predictions and we also need to select the
machine-learning algorithm we want to use to generate patient-level predictions.
Currently, the available options are regularized logistic regression, random
forest, gradient boosting machines, decision tree, naive Bayes, K-nearest
neighbors, neural network and deep learning (convolutional neural networks,
recurrent neural network and deep nets).

@@

The number of patients and person years by treatment arm along with the number
of outcomes should be reported. A performance overview of the derived prediction
models, including discrimination and calibration both in the propensity score
matched subset, the entire population and separately for treated and comparator
patients should also be reported. This is quite relevant as the performance of
the prediction models is directly related to our ability to single out patient
subgroups where treatment may be highly beneficial or unsafe. Kent et al
[@Kent2016] demonstrated that the event rate and the discriminative ability of
the prediction model can predict very well the distribution of predicted risk.
Lower event rate and higher c-statistic (given good calibration) result in high
risk heterogeneity, thus making estimated average treatment effects
uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

In every risk stratum propensity score distributions by treatment group and
covariate balance plots should be made available along with event rates, hazard
ratios and absolute risk differences. Finally, the results should be generated
with a standardized format to allow for combination of results from multiple
data sources. This also enables the development of scalable demonstration tools
such as R-shiny applications, interactive tables and figures.

@@

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a useful overview
for clinical interpretation, it is not appropriate to guide clinical practice,
where decisions need to be made at the individual level [@Kent2019].
Presentation of treatment effects as a continuous function of risk would be more
helpful, but is methodologically challenging. Future research is necessary for
the development of methods for continuous risk-based assessment of HTE.

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014]. However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, this may introduce spurious
treatment-covariate interactions in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.
